---
title: "Exploring the gapminder using dplyr and ggplot2"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document, we will be exploring the `gapminder` dataset using (among other tools) the `dplyr` and `ggplot2` functions. First we load the gapminder dataset and dplyr using the tidyverse package.
```{r}
library(gapminder)
library(tidyverse)
```
##Smell test the data
In this section we determine some basic attributes of `gapminder`.

 - Is it a data.frame, a matrix, a vector, a list?
```{r}
typeof(gapminder)
head(gapminder)
```
 
 
 - What is its class?
```{r}
class(gapminder)
```
 
 - How many variables/columns?
```{r}
ncol(gapminder)
```
 
 
 - How many rows/observations?
 
```{r}
nrow(gapminder)
```
 
 - Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different contexts?
 
We can also determine both the class and the number of rows & columns (among other information) by using the `str` function,
```{r}
str(gapminder)
```
 which also displays the class(es) of gapminder and the number of rows & columns.
 
 - What data type is each variable?
 
 From above, we see that the continent and country variables are characters, year and population are integers, and life expectancy and gdp per capita are numerical (double).

 
 

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
