---
title: "Exploring the gapminder using dplyr and ggplot2"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document, we will be exploring the `gapminder` dataset using (among other tools) the `dplyr` and `ggplot2` functions. First we load the gapminder dataset and dplyr using the tidyverse package.
```{r}
library(gapminder)
library(tidyverse)
```
##Smell test the data
In this section we determine some basic attributes of `gapminder`.

 - Is it a data.frame, a matrix, a vector, a list?
```{r}
typeof(gapminder)
head(gapminder)
summary(gapminder)
typeof(gapminder)
```
 
 
 - What is its class?
```{r}
class(gapminder)
```
 
 - How many variables/columns?
```{r}
ncol(gapminder)
```
 
 
 - How many rows/observations?
 
```{r}
nrow(gapminder)
```
 
 - Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different contexts?
 
 The `ncol` and `nrow` functions return the number of columns and rows, respectively, of `gapminder`. We can also extract the same information by using the `str` function,
```{r}
str(gapminder)
```
 which also displays the class of gapminder.
 
 - What data type is each variable?
```{r}

```
 
 

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
